# Linear regression models: Diagnostics and model-building

```{r}
pacman::p_load(tidyverse, GLMsData, janitor, patchwork)
```

```{r}
data("lungcap")
lung <- lungcap |>
  as_tibble() |>
  clean_names() |>
  mutate(smoke = factor(
    smoke,
    levels = c(0, 1),
    labels = c('Non-smoker',
               'Smoker')))
```

## Residuals for non-normal linear models

```{r}
model <- lung |> lm(fev ~ ht + gender + smoke, data = _)

c('Residuals' = var(resid(model)), 
  'Stadarized residuals' = var(rstandard(model)))
```

```{r}
p1 <- 
  resid(model) |> 
  as_tibble() |> 
  ggplot(aes(value)) +
  geom_histogram(bins = 50)

p2 <- 
  rstandard(model) |> 
  as_tibble() |> 
  ggplot(aes(value)) +
  geom_histogram(bins = 50)

p1 + p2
```

## Leverages for linear regression models

```{r}
hat_values <- hatvalues(model)

# Two largest leverages
sort(hat_values, decreasing = TRUE)[1:2]

# Mean of leverages
mean(hat_values); length(coef(model)) / length(lung$fev)
sort(hat_values, decreasing = TRUE)[1:2] / mean(hat_values)
```

```{r}
sort_h <- sort(hat_values, decreasing = TRUE, index.return = TRUE)
large_h <- sort_h$ix[1:2]
lung[large_h, ]
```

```{r}
lung |> 
  rowid_to_column() |> 
  mutate(color = case_when(
    rowid %in% large_h ~ 1,
    TRUE ~ 0
  )) |> 
  filter(gender == 'M' & smoke == 'Smoker') |> 
  ggplot(aes(ht, fev, color = color)) +
  geom_point() +
  theme(legend.position = 'none') +
  ggtitle('Male smokers')
```

## Residual plots

### Residuals against $x_j$

```{r}
scatter.smooth(rstandard(model) ~ lung$ht, col = 'gray')
```

```{r}
bind_cols(lung, rstandard(model) |> as_tibble()) |> 
  ggplot(aes(ht, value)) +
  geom_point(alpha = .5) +
  geom_smooth(method = 'loess', formula = 'y ~ x')
```

### Partial residual plots

```{r}
partial_resid <- resid(model, type = 'partial') |> as_tibble(); partial_resid
```

```{r}
termplot(model, partial.resid = TRUE, terms = 'ht', las = 1)
```

```{r}
bind_cols(lung, partial_resid |> select(partial_resid = ht)) |> 
  ggplot(aes(ht, partial_resid)) +
  geom_point(alpha = .5) +
  geom_smooth(method = 'loess', formula = 'y ~ x')
```

```{r}
coef(summary(model))

ht_model <- 
  bind_cols(lung, partial_resid |> select(partial_resid = ht)) |> 
  lm(partial_resid ~ ht, data = _)

coef(summary(ht_model))
```

### Plot residuals against $\hat{\mu}$: Constant variance

```{r}
scatter.smooth(rstandard(model) ~ fitted(model), col = 'grey')
```

```{r}
bind_cols(rstandard(model), fitted(model)) |> 
  rename(resid = 1, fitted = 2) |> 
  ggplot(aes(fitted, resid)) +
  geom_point(alpha = .5) +
  geom_smooth(method = 'loess', formula = 'y ~ x')
```

### Q---Q plots and normality

```{r}
qqnorm(rstandard(model))
qqline(rstandard(model))
```

```{r}
resid <- rstandard(model) |> as_tibble()

p1 <- 
  resid |> 
  ggplot(aes(sample = value)) +
  stat_qq() +
  stat_qq_line()

p2 <- 
  resid |> 
  ggplot(aes(value)) +
  geom_histogram(bins = 50)

p1 + p2
```

## Outliers and influential observations

### Studentized residuals

```{r}
summary(cbind(Standarized = rstandard(model), Studentized = rstudent(model)))
```

### Influential observations

```{r}
model <- lung |> lm(fev ~ ht + gender + smoke, data = _)
```

```{r}
# Cook' distance
cooks_max <- which.max(cooks.distance(model))
cooks_min <- which.min(cooks.distance(model))
c(Min = cooks_min, Max = cooks_max)

# DFFITS, CV and Cook's distance
out <- cbind(
  DFFITS = dffits(model),
  Cooks = cooks.distance(model),
  Cov_rat = covratio(model)
)

round(out[c(cooks_min, cooks_max), ], 5)
```

```{r}
model |> broom::tidy()
dfbetas(model)[cooks_min, ]
dfbetas(model)[cooks_max, ]
```

```{r}
influence <- influence.measures(model); names(influence)

head(round(influence$infmat, 3))
head(influence$is.inf)
```

```{r}
colSums(influence$is.inf)

# 7 observations have high leverage
# 56 observations are identified by the coveriance ratio as influential
# Cook's distance does not identify any observation as influential
```

```{r}
table(rowSums(influence$is.inf[, -8]))

# 54 observations are declared as influential on just one criterion
```

#### Influence diagnostics

```{r}
bind_cols(
  cooks.distance(model) |> as_tibble_col('cooks'),
  dffits(model) |> as_tibble_col('dffits'),
  dfbetas(model) |> as_tibble() |> select(3)
) |> 
  rowid_to_column() |> 
  pivot_longer(!rowid) |> 
  ggplot(aes(rowid, value, color = name)) +
  geom_line() +
  facet_wrap(~ name, scales = 'free_y')
```

## Transforming the response

```{r}
model_sqrt <- lung |> lm(sqrt(fev) ~ ht + gender + smoke, data = _)

bind_cols(
  rstandard(model_sqrt) |> as_tibble_col('standardized'),
  fitted(model_sqrt) |> as_tibble_col('fitted')
) |> 
  ggplot(aes(fitted, standardized)) +
  geom_point()
```

```{r}
model_log <- lung |> lm(log(fev) ~ ht + gender + smoke, data = _)

bind_cols(
  rstandard(model_log) |> as_tibble_col('log'),
  fitted(model_log) |> as_tibble_col('fitted')
) |> 
  ggplot(aes(fitted, log)) +
  geom_point()
```

#### The Box-Cox transformation

```{r}
library(MASS)

boxcox(fev ~ ht + gender + smoke,
       lambda = seq(-.25, .25, length = 11),
       data = lung)
```
